---
title: "PHQ-9 DIF Analysis"
output: html_document
date: "2024-09-01"
---

```{r, include = F}
## Set seed
set.seed(57135)


## Load packages
library(here)
library(lordif)
library(ltm)
library(mirt)
library(kableExtra)
library(broom)
library(scales)
library(tidyverse)
`%+%` <- paste0


## Load data
df <- readRDS(here("Data", "Clean Data.rds"))


## Custom functions
pull_scaled_theta <- function(model) {
  
  model %>%
    fscores() %>%
    scale() %>%
    as.numeric() %>%
    return()
  
}

plot_mean_response <- function(model, theta = "theta_fcm") {
  
  class <- class(model)
  
  min_theta <- min(df[[theta]])
  max_theta <- max(df[[theta]])
  
  out <- expand_grid(
    theta_temp = seq(min_theta, max_theta, .01),
    sample_hms = c(0, 1)
  )
  
  names(out)[names(out) == "theta_temp"] <- theta
  
  predictions <- predict(
    object = model,
    newdata = out,
    type = if_else(
      "lm" %in% class,
      "response",
      "mean"
    )
  )
  
  out$pred <- predictions
  
  plot <- out %>%
    mutate(
      sample_hms = case_match(
        sample_hms, 
        0 ~ "NHANES", 
        1 ~ "HMS" 
      )
    ) %>%
    ggplot() +
    geom_line(
      aes(.data[[theta]], pred, group = sample_hms, color = sample_hms),
      linewidth = 1
    ) +
    scale_x_continuous(
      name = expression(paste("Estimated Depression (", theta, ")"))
    ) +
    scale_color_discrete(
      name = "Sample"
    ) +
    theme_classic()
  
  if(any("lrm" %in% class)) {
    
    plot <- plot + 
        scale_y_continuous(
        name = "Predicted Response",
        limits = c(0, 3),
        breaks = 0:3,
        labels = c("Not at all", "Several days", "More than half the days", "Nearly every day")
      )
    
  }
  
  return(plot)
  
}

plot_response_curves <- function(model, theta = "theta_fcm") {
  
  min_theta <- min(df[[theta]])
  max_theta <- max(df[[theta]])
  
  out <- expand_grid(
    theta_temp = seq(min_theta, max_theta, .01),
    sample_hms = c(0, 1)
  )
  
  names(out)[names(out) == "theta_temp"] <- theta
  
  predictions <- predict(
    object = model,
    newdata = out,
    type = "fitted.ind"
  )
  
  out$pred_0 <- as.numeric(predictions[, 1])
  out$pred_1 <- as.numeric(predictions[, 2])
  out$pred_2 <- as.numeric(predictions[, 3])
  out$pred_3 <- as.numeric(predictions[, 4])
  
  plot <- out %>%
    mutate(
      sample_hms = case_match(
        sample_hms, 
        0 ~ "NHANES", 
        1 ~ "HMS" 
      )
    ) %>%
    pivot_longer(
      cols = c("pred_0", "pred_1", "pred_2", "pred_3"),
      names_to = "response_level",
      values_to = "prob"
    ) %>%
    ggplot() +
    geom_line(
      aes(.data[[theta]], prob, group = response_level, color = response_level),
      linewidth = 1
    ) +
    scale_x_continuous(
      name = expression(paste("Estimated Depression (", theta, ")"))
    ) +
    scale_y_continuous(
      name = "Probability"
    ) +
    scale_color_discrete(
      name = "Response Level",
      labels = c("Not at all", "Several days", "More than half the days", "Nearly every day")
    ) +
    facet_wrap(~ sample_hms, ncol = 1) +
    theme_classic()
  
  return(plot)
  
}

or_lrm <- function(model) {
  
  coef <- model$coefficients
  
  se <- model$var %>%
    diag() %>%
    sqrt()
  
  out <- tibble(
    var = names(coef),
    or = exp(coef),
    high = exp(coef + (se * qnorm(.975))),
    low = exp(coef + (se * qnorm(.025)))
  )
  
  return(out)
  
}

tidy_lrm <- function(model) {
  
  r2 <- tibble(
    var = "R2",
    or = model$stats[["R2"]] %>%
      round(3) %>%
      as.character()
  )
  
  model %>%
    or_lrm() %>%
    mutate(
      or = paste0(
        round(or, 2),
        " (",
        round(low, 2),
        ", ",
        round(high, 2),
        ")"
      )
    ) %>%
    select(var, or) %>%
    bind_rows(r2) %>%
    return()
  
}

tidy_lrm_list <- function(model_list, theta = "theta_fcm") {
  
  out <- tibble(
    var = c("y>=1", "y>=2", "y>=3", theta, "sample_hms", theta %+% " * sample_hms", "R2")
  )
  
  tidy_output_1 <- tidy_lrm(
    model_list[[1]]
  )
  
  names(tidy_output_1)[names(tidy_output_1) == "or"] <- "or_1"
    
  out <- left_join(out, tidy_output_1, by = "var")
  
  for(i in 2:length(model_list)) {
    
    tidy_output_i <- tidy_lrm(
      model_list[[i]]
    )
    
    names(tidy_output_i)[names(tidy_output_i) == "or"] <- "or_" %+% i
    
    out <- left_join(out, tidy_output_i, by = "var")
    
  }
  
  return(out)
  
}

release_and_test <- function(item, theta = "theta_fcm") {
  
  released_model <- multipleGroup(
    data = df[, 1:9],
    model = 1,
    group = df$sample_char,
    invariance = setdiff("phq_" %+% 1:9, item),
    verbose = F
  )
  
  theta_released <- pull_scaled_theta(released_model)
  
  print("Correlation between released theta and original theta: " %+% round(cor(df[[theta]], theta_released), 2))
  print("Proportion of participants whose theta changes more than .3 units: " %+% percent(mean((df[[theta]] - theta_released) > .3), .1))
  
}
```

## Global DIF Test

```{r, include = F}
# Calculate DIF via lordif
dif <- lordif(
  resp.data = data.frame(df),
  group = df$sample_hms,
  selection = 1:9,
  criterion = "R2",
  pseudo.R2 = "Nagelkerke",
  R2.change = .02
  # weights = df$scaled_weight
)
```

```{r}
# Table 1
dif %>%
  summary() %>%
  pluck("stats") %>%
  select(item, chi12, chi23, beta12, pseudo12.Nagelkerke, pseudo23.Nagelkerke) %>%
  mutate(
    across(c(chi12, chi23), ~ round(., 3)),
    beta12 = percent(beta12, .1),
    across(c(pseudo12.Nagelkerke, pseudo23.Nagelkerke), ~ round(., 3))
  ) %>%
  kable() %>%
  kable_styling()

# Pull theta
df$theta_lordif <- as.numeric(dif$calib$theta)

# In the absence of DIF, calculate fully constrained IRT model; thetas should match
fully_constrained_model <- multipleGroup(
  data = df[, 1:9],
  model = 1,
  group = df$sample_char,
  invariance = "phq_" %+% 1:9,
  method = "EM", # "Standard EM algorithm with fixed quadrature"
  verbose = F
)

# coef(fully_constrained_model)

# Pull theta
df$theta_fcm <- pull_scaled_theta(fully_constrained_model)

# Correlations between thetas, phq_sum
df %>%
  select(phq_sum, theta_lordif, theta_fcm) %>%
  cor() %>%
  kable() %>%
  kable_styling()
```

### Item 1: Anhedonia

```{r}
# Calculate models
phq_1_model_1 <- lrm(data = df, formula = phq_1 ~ theta_fcm)
phq_1_model_2 <- lrm(data = df, formula = phq_1 ~ theta_fcm + sample_hms)
phq_1_model_3 <- lrm(data = df, formula = phq_1 ~ theta_fcm * sample_hms)

# Print models
tidy_lrm_list(list(phq_1_model_1, phq_1_model_2, phq_1_model_3), "theta_fcm") %>%
  kable() %>%
  kable_styling()

# Print plots
plot_mean_response(phq_1_model_3) +
  ggtitle(
    "Predicted Item Response by Theta and Group",
    "PHQ-9 Item 1: Anhedonia"
  )

plot_response_curves(phq_1_model_3) +
  ggtitle(
    "Predicted Item Response by Theta and Group",
    "PHQ-9 Item 1: Anhedonia"
  )

# Measure change in theta in constrained model
release_and_test("phq_1")
```

### Item 2: Depressed Mood

```{r}
# Calculate models
phq_2_model_1 <- lrm(data = df, formula = phq_2 ~ theta_fcm)
phq_2_model_2 <- lrm(data = df, formula = phq_2 ~ theta_fcm + sample_hms)
phq_2_model_3 <- lrm(data = df, formula = phq_2 ~ theta_fcm * sample_hms)

# Print models
tidy_lrm_list(list(phq_2_model_1, phq_2_model_2, phq_2_model_3)) %>%
  kable() %>%
  kable_styling()

# Print plots
plot_mean_response(phq_2_model_3) +
  ggtitle(
    "Predicted Item Response by Theta and Group",
    "PHQ-9 Item 2: Depressed Mood"
  )

plot_response_curves(phq_2_model_3) +
  ggtitle(
    "Predicted Item Response by Theta and Group",
    "PHQ-9 Item 2: Depressed Mood"
  )

# Trying constrained model, per Paul's email
release_and_test("phq_2")
```

### Item 3: Trouble Sleeping

```{r}
# Calculate models
phq_3_model_1 <- lrm(data = df, formula = phq_3 ~ theta_fcm)
phq_3_model_2 <- lrm(data = df, formula = phq_3 ~ theta_fcm + sample_hms)
phq_3_model_3 <- lrm(data = df, formula = phq_3 ~ theta_fcm * sample_hms)

# Print models
tidy_lrm_list(list(phq_3_model_1, phq_3_model_2, phq_3_model_3)) %>%
  kable() %>%
  kable_styling()

# Print plots
plot_mean_response(phq_3_model_3) +
  ggtitle(
    "Predicted Item Response by Theta and Group",
    "PHQ-9 Item 3: Trouble Sleeping"
  )

plot_response_curves(phq_3_model_3) +
  ggtitle(
    "Predicted Item Response by Theta and Group",
    "PHQ-9 Item 3: Trouble Sleeping"
  )

# Trying constrained model, per Paul's email
release_and_test("phq_3")
```

### Item 4: Feeling Tired

```{r}
# Calculate models
phq_4_model_1 <- lrm(data = df, formula = phq_4 ~ theta_fcm)
phq_4_model_2 <- lrm(data = df, formula = phq_4 ~ theta_fcm + sample_hms)
phq_4_model_3 <- lrm(data = df, formula = phq_4 ~ theta_fcm * sample_hms)

# Print models
tidy_lrm_list(list(phq_4_model_1, phq_4_model_2, phq_4_model_3)) %>%
  kable() %>%
  kable_styling()

# Print plots
plot_mean_response(phq_4_model_3) +
  ggtitle(
    "Predicted Item Response by Theta and Group",
    "PHQ-9 Item 4: Feeling Tired"
  )

plot_response_curves(phq_4_model_3) +
  ggtitle(
    "Predicted Item Response by Theta and Group",
    "PHQ-9 Item 4: Feeling Tired"
  )

# Trying constrained model, per Paul's email
release_and_test("phq_4")
```

### Item 5: Poor Appetite or Overeating

```{r}
# Calculate models
phq_5_model_1 <- lrm(data = df, formula = phq_5 ~ theta_fcm)
phq_5_model_2 <- lrm(data = df, formula = phq_5 ~ theta_fcm + sample_hms)
phq_5_model_3 <- lrm(data = df, formula = phq_5 ~ theta_fcm * sample_hms)

# Print models
tidy_lrm_list(list(phq_5_model_1, phq_5_model_2, phq_5_model_3)) %>%
  kable() %>%
  kable_styling()

# Print plots
plot_mean_response(phq_5_model_3) +
  ggtitle(
    "Predicted Item Response by Theta and Group",
    "PHQ-9 Item 5: Poor Appetite or Overeating"
  )

plot_response_curves(phq_5_model_3) +
  ggtitle(
    "Predicted Item Response by Theta and Group",
    "PHQ-9 Item 5: Poor Appetite or Overeating"
  )

# Trying constrained model, per Paul's email
release_and_test("phq_5")
```

### Item 6: Feeling Bad About Yourself

```{r}
# Calculate models
phq_6_model_1 <- lrm(data = df, formula = phq_6 ~ theta_fcm)
phq_6_model_2 <- lrm(data = df, formula = phq_6 ~ theta_fcm + sample_hms)
phq_6_model_3 <- lrm(data = df, formula = phq_6 ~ theta_fcm * sample_hms)

# Print models
tidy_lrm_list(list(phq_6_model_1, phq_6_model_2, phq_6_model_3)) %>%
  kable() %>%
  kable_styling()

# Print plots
plot_mean_response(phq_6_model_3) +
  ggtitle(
    "Predicted Item Response by Theta and Group",
    "PHQ-9 Item 6: Feeling Bad About Yourself"
  )

plot_response_curves(phq_6_model_3) +
  ggtitle(
    "Predicted Item Response by Theta and Group",
    "PHQ-9 Item 6: Feeling Bad About Yourself"
  )

# Trying constrained model, per Paul's email
release_and_test("phq_6")
```

### Item 7: Difficulty Concentrating

```{r}
# Calculate models
phq_7_model_1 <- lrm(data = df, formula = phq_7 ~ theta_fcm)
phq_7_model_2 <- lrm(data = df, formula = phq_7 ~ theta_fcm + sample_hms)
phq_7_model_3 <- lrm(data = df, formula = phq_7 ~ theta_fcm * sample_hms)

# Print models
tidy_lrm_list(list(phq_7_model_1, phq_7_model_2, phq_7_model_3)) %>%
  kable() %>%
  kable_styling()

# Print plots
plot_mean_response(phq_7_model_3) +
  ggtitle(
    "Predicted Item Response by Theta and Group",
    "PHQ-9 Item 7: Difficulty Concentrating"
  )

plot_response_curves(phq_7_model_3) +
  ggtitle(
    "Predicted Item Response by Theta and Group",
    "PHQ-9 Item 7: Difficulty Concentrating"
  )

# Trying constrained model, per Paul's email
release_and_test("phq_7")
```

### Item 8: Psychomotor Symptoms

```{r}
# Calculate models
phq_8_model_1 <- lrm(data = df, formula = phq_8 ~ theta_fcm)
phq_8_model_2 <- lrm(data = df, formula = phq_8 ~ theta_fcm + sample_hms)
phq_8_model_3 <- lrm(data = df, formula = phq_8 ~ theta_fcm * sample_hms)

# Print models
tidy_lrm_list(list(phq_8_model_1, phq_8_model_2, phq_8_model_3)) %>%
  kable() %>%
  kable_styling()

# Print plots
plot_mean_response(phq_8_model_3) +
  ggtitle(
    "Predicted Item Response by Theta and Group",
    "PHQ-9 Item 8: Psychomotor Symptoms"
  )

plot_response_curves(phq_8_model_3) +
  ggtitle(
    "Predicted Item Response by Theta and Group",
    "PHQ-9 Item 8: Psychomotor Symptoms"
  )

# Trying constrained model, per Paul's email
release_and_test("phq_8")
```

### Item 9: Suicidal Ideation

```{r}
# Calculate models
phq_9_model_1 <- lrm(data = df, formula = phq_9 ~ theta_fcm)
phq_9_model_2 <- lrm(data = df, formula = phq_9 ~ theta_fcm + sample_hms)
phq_9_model_3 <- lrm(data = df, formula = phq_9 ~ theta_fcm * sample_hms)

# Print models
tidy_lrm_list(list(phq_9_model_1, phq_9_model_2, phq_9_model_3)) %>%
  kable() %>%
  kable_styling()

# Print plots
plot_mean_response(phq_9_model_3) +
  ggtitle(
    "Predicted Item Response by Theta and Group",
    "PHQ-9 Item 9: Suicidal Ideation"
  )

plot_response_curves(phq_9_model_3) +
  ggtitle(
    "Predicted Item Response by Theta and Group",
    "PHQ-9 Item 9: Suicidal Ideation"
  )

# Trying constrained model, per Paul's email
release_and_test("phq_9")
```

### Combined plot

```{r}
min_theta <- min(df$theta_fcm)
max_theta <- max(df$theta_fcm)

out <- expand_grid(
  theta_fcm = seq(min_theta, max_theta, .01),
  sample_hms = c(0, 1)
)

predictions <- list(
  phq_1_model_3,
  phq_2_model_3,
  phq_3_model_3,
  phq_4_model_3,
  phq_5_model_3,
  phq_6_model_3,
  phq_7_model_3,
  phq_8_model_3,
  phq_9_model_3
) %>%
  map(
    .,
    ~ predict(
      object = .,
      newdata = out,
      type = "mean"
    )
  ) %>%
  bind_cols() %>%
  rename(
    `Item 1: Anhedonia` = ...1,
    `Item 2: Depressed Mood` = ...2,
    `Item 3: Trouble Sleeping` = ...3,
    `Item 4: Feeling Tired` = ...4,
    `Item 5: Poor Appetite or Overeating` = ...5,
    `Item 6: Feeling Bad About Yourself` = ...6,
    `Item 7: Difficulty Concentrating` = ...7,
    `Item 8: Psychomotor Symptoms` = ...8,
    `Item 9: Suicidal Ideation` = ...9
  )

figure_1 <- out %>%
  bind_cols(predictions) %>%
  pivot_longer(
    cols = starts_with("Item"),
    names_to = "item",
    values_to = "pred"
  ) %>%
  mutate(
    sample_hms = case_match(
      sample_hms, 
      0 ~ "NHANES", 
      1 ~ "HMS" 
    )
  ) %>%
  ggplot() +
  geom_line(
    aes(theta_fcm, pred, group = sample_hms, color = sample_hms),
    linewidth = 1
  ) +
  scale_x_continuous(
    name = expression(paste("Estimated Depression (", theta, ")"))
  ) +
  scale_y_continuous(
    name = "Predicted Response",
    limits = c(0, 3),
    breaks = 0:3,
    labels = c("Not at all", "Several days", "More than half the days", "Nearly every day")
  ) +
  scale_color_discrete(
    name = "Sample"
  ) +
  facet_wrap(
    ~ item,
    ncol = 3,
    nrow = 3
  ) +
  theme_classic() +
  ggtitle("Predicted Item Response by Theta and Group")

figure_1

ggsave(
  filename = here("Figure 1.png"),
  plot = figure_1,
  width = 10,
  height = 8,
  units = "in"
)
```

### Combined Table

```{r}
tidy_lrm_list(
  list(
    phq_1_model_2,
    phq_2_model_2,
    phq_3_model_2,
    phq_4_model_2,
    phq_5_model_2,
    phq_6_model_2,
    phq_7_model_2,
    phq_8_model_2,
    phq_9_model_2
  )
) %>%
  drop_na() %>%
  kable() %>%
  kable_styling()
```

### Sum Score

Total scores correspond very closely.

```{r}
# Linear model predicting sum score
sum_score_model <- lm(data = df, formula = phq_sum ~ poly(scale(theta_fcm), 3) + sample_hms)
sum_score_model %>%
  tidy() %>%
  kable(digits = 2) %>%
  kable_styling()

# Plot using smoother
df %>%
  ggplot() +
  geom_smooth(
    aes(theta_fcm, phq_sum, group = sample_char, color = sample_char),
    method = "gam", # Generalized additive model via mgcv::gam()
    formula = y ~ s(x, bs = "cs"),
    se = F
  ) +
  scale_x_continuous(name = expression(paste("Estimated Depression (", theta, ")"))) +
  scale_y_continuous(name = "PHQ-9 Sum Score") +
  scale_color_discrete(name = "Sample") +
  ggtitle("Predicted PHQ-9 Sum Score by Theta and Group") +
  theme_classic()

# Logistic models predicting categorical cutoff don't seem to fit well, as quadratic
# terms resulted in predicted values of 1 or 0 (probably because of high R2)... 
# but based on the geom_smooth() above I think it's safe to say this isn't a problem
```

### Exploratory: What if we hold the core symptoms constant, but allow all others to vary?

```{r}
core_symptom_model <- multipleGroup(
  data = df[, 1:9],
  model = 1,
  group = df$sample_char,
  invariance = c("phq_1", "phq_2"),
  verbose = F
)

# coef(core_symptom_model)

df$theta_csm <- pull_scaled_theta(core_symptom_model)

# Correlations between thetas across models
df %>%
  select(starts_with("theta")) %>%
  cor() %>%
  kable() %>%
  kable_styling()

# Differences in this model
df %>%
  mutate(theta_change = theta_csm - theta_fcm) %>%
  group_by(sample_char) %>%
  summarize(
    mean_change = mean(theta_change),
    significant_change = mean(theta_change > .3)
  ) %>%
  kable() %>%
  kable_styling()

# Logistic models
"phq_" %+% 1:9 %>%
  map(.,
      ~ lrm(
        data = df,
        formula = as.formula(. %+% " ~ theta_csm + sample_hms")
      )) %>%
  tidy_lrm_list(theta = "theta_csm") %>%
  drop_na() %>%
  kable() %>%
  kable_styling()
```

### Exploratory: What if we use PHQ-2 sum score as our "theta"?

```{r}
df$core_symptom_sum <- df$phq_1 + df$phq_2

# Correlations between sum scores and thetas
df %>%
  select(core_symptom_sum, phq_sum, starts_with("theta")) %>%
  cor() %>%
  kable() %>%
  kable_styling()

# Logistic models
"phq_" %+% 1:9 %>%
  map(.,
      ~ lrm(
        data = df,
        formula = as.formula(. %+% " ~ core_symptom_sum + sample_hms")
      )) %>%
  tidy_lrm_list(theta = "core_symptom_sum") %>%
  drop_na() %>%
  kable() %>%
  kable_styling()
```

Do we get different estimates of depression using PHQ-9 and PHQ-2 cutoffs?

```{r}
library(survey)

df <- df %>%
  mutate(
    across(
      all_of("phq_" %+% 1:9),
      ~ . >= 2,
      .names = "{.col}_bin"
    ),
    elevated_symptoms_total = phq_1_bin + phq_2_bin + phq_3_bin + phq_4_bin + phq_5_bin + phq_6_bin + phq_7_bin + phq_8_bin + phq_9_bin,
    elevated_core_symptoms = phq_1_bin + phq_2_bin,
    diagnostic_algorithm = elevated_symptoms_total >= 5 & elevated_core_symptoms >= 1
  )

df_weights <- df %>%
  drop_na(weight)

design <- svydesign(
  data = df_weights,
  ids = ~ 0,
  strata = ~ sample_char,
  weights = ~ weight
)

svyby(~ phq_sum >= 10, ~ sample_char, design, svyciprop, vartype = "ci") # Good properties vs diagnostic interviews (1)
svyby(~ core_symptom_sum >= 3, ~ sample_char, design, svyciprop, vartype = "ci") # Good properties (2)
svyby(~ core_symptom_sum >= 3 & phq_sum >= 10, ~ sample_char, design, svyciprop, vartype = "ci") # Even better properties (2) 
svyby(~ diagnostic_algorithm, ~ sample_char, design, svyciprop, vartype = "ci") # Not recommended (1) (3)

# (1) https://karger.com/pps/article/89/1/25/289799/The-Accuracy-of-the-Patient-Health-Questionnaire-9
# (2) https://jamanetwork.com/journals/jama/article-abstract/2766865
# (3) https://www.sciencedirect.com/science/article/abs/pii/S0163834314002540?via%3Dihub
```